---
title: "R"
author: "Africano"
date: "2023-03-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Create workspace

1.  Create project to a given folder
2.  Get the required resources to the folder (data)
3.  Create and save an RMD that you will work in

# install the required R packages

```{r install, eval=FALSE}

install.packages("readr") 

install.packages("dplyr")

install.packages("gtsummary")

install.packages("kablExtra")

# OR

install.packages(c("readr", "dplyr")) # install multiple packages
```

Note:

1.  you can install using the panes there fom GUI of R studio install in Packages panel (refer to our note for instructions)

2.  You only need to install the package once, call it everytime when you want to use it.

# import data

\* use command (depend on file type)

```{r}
BirdNest <- readr::read_csv("BirdNest.csv")  # read data from csv file
```

# Introduction to BirdNest

Nest and species characteristics for North American passerines

Amy R. Moore, as a student at Grinnell College in 1999, wanted to study the relationship between species characteristics and the type of nest a bird builds, using data collected from available sources. For the study, she collected data by species for 84 separate species of North American passerines. **It is A data frame with 84 observations on the following 12 variables as shown below:**

| **`Species`**  | Latin species name                                                                     |
|----------------|----------------------------------------------------------------------------------------|
| **`Common`**   | Common species name                                                                    |
| **`Page`**     | Page in a bird manual describing the species                                           |
| **`Length`**   | Mean body length for the species (in cm)                                               |
| **`Nesttype`** | Type of nest                                                                           |
| **`Location`** | Location of nest                                                                       |
| **`No.eggs`**  | Number of eggs                                                                         |
| **`Color`**    | Egg color (**`0`**=plain/solid or **`1`**=speckled/spotted)                            |
| **`Incubate`** | Mean length of time (in days) the species incubates eggs in the nest                   |
| **`Nestling`** | Mean length of time (in days) the species cares for babies in the nest until fledged   |
| **`Totcare`**  | Total care time = Incubate+Nestling                                                    |
| **`Closed`**   | 1=closed nest (pendant, spherical, cavity, crevice, burrow), 0=open nest (saucer, cup) |

# Basic functions to learn about your data

```{r}

summary(BirdNest) # describe data
```

```{r}
sapply(BirdNest, class) # check the type of each variable in BirdNest
```

```{r}
dim(BirdNest)  # dimension
colnames(BirdNest) # column names
```

```{r}
BirdNest[!complete.cases(BirdNest),] # return all rows with missing value in BirdNest; use "complete.cases()" for non-missings; use "!" for negation (NOT)
```

```{r}
class(BirdNest)  # the type of the data set 
```

```{r}
head(BirdNest)  # first six rows
```

```{r}
BirdNest$Location %>% 
  table()
# return all unique value with frequency in Location variable of BirdNest

# Or, for a better output
library(gtsummary)

BirdNest$Location %>% 
  tbl_summary()

```

```{r}
unique(BirdNest$Location) # return all unique value in factor
```

```{r}
# return the number of unique value in factor
BirdNest$Location %>% 
  unique() %>% 
  length()

```

```{r}
BirdNest$Location %>% 
  is.na() %>% 
  table() %>% 

BirdNest$Totcare %>% 
  is.na() %>% 
  table()

```

# Data Wrangling

## select

To select columns or drop columns of a data frame, use select()

### Select desired variables

```{r}
#Select four columns: Length, Nesttype, Location, No.eggs from original data, return the first six rows.
library(dplyr)
BirdNest %>% 
  select(Length, Nesttype, Location, No.eggs) %>% 
  head()
```

### Drop undesired variables

```{r}
#Remove two columns: Species and Common from original data  
BirdNest %>% 
  select(-c(Species,Common)) %>% 
  head()
```

**Note: You may want to set some criteria for variable inclusion when you have a lot of variables in your data.**

+--------------+--------------------------------------------------------+
| usage        | summary                                                |
+:=============+:=======================================================+
| -            | Select everything but                                  |
+--------------+--------------------------------------------------------+
| :            | Select range                                           |
+--------------+--------------------------------------------------------+
| contains()   | Select columns whose name contains a character string  |
+--------------+--------------------------------------------------------+
| start_with() | Select columns whose name starts with a string         |
+--------------+--------------------------------------------------------+
| ends_with()  | Select columns whose name ends with a string           |
+--------------+--------------------------------------------------------+
| matches()    | Select columns whose name matches a regular expression |
+--------------+--------------------------------------------------------+
| one_of()     | Select columns whose names are in a group of names     |
+--------------+--------------------------------------------------------+

```{r}
# Select variables contain "nest" (default case insensitive)
BirdNest %>% 
  select(contains("nest")) %>% 
  head()
```

### filter

Use **`filter`** to select rows that meet criteria.

```{r}
library(dplyr)
#Select rows with Length more than 30
BirdNest %>% 
  filter(Length > 30)

#Select rows with No.eggs more than 6 and Location in "decid" 
BirdNest %>% 
  filter(No.eggs > 6 & Location == "decid")

```

**Try Out:** Could you filter the rows with levels contains "Jay" in Common column? (hint: use grep function)

```{r}
BirdNest %>% 
  filter(grepl("Jay",Common))

```

### mutate

When you want to create new columns based on the values in existing columns, for example, do calculation using existing variables, we'll use the dplyr function mutate().

```{r}

head(mutate(BirdNest, ratio = Incubate/Totcare))
head(mutate(BirdNest, ratio = Incubate/Totcare, inverse = 1/ratio))
head(mutate(BirdNest, cumsum_total = cumsum(Totcare)))
head(mutate(BirdNest, nor_Nest = Nestling/mean(Nestling, na.rm=T))) # na.rm=T removes the missing value when calculating mean 
```

```{r}
BirdNest %>% 
  mutate(ratio = Incubate/Totcare) %>% 
  head()
```

```{r}
BirdNest %>% 
  mutate(ratio = Incubate/Totcare, inverse = 1/ratio) %>% 
  head()
```

```{r}
BirdNest %>% 
  mutate(cumsum_total = cumsum(Totcare)) %>% 
  head()
```

```{r}
BirdNest %>% 
  mutate(nor_Nest = Nestling/mean(Nestling, na.rm=T)) %>% 
  head()
# na.rm=T removes the missing value when calculating mean 
```

**Note: You can create variables with criteria**

| usage               | summary                        |
|:--------------------|:-------------------------------|
| pmin(), pmax()      | Elementwise minimum or maximum |
| cummin(), cummax()  | Cumulative minimum and maximum |
| cumsum(), cumprod() | Cumulative sum and product     |

### arrange

Arrange the rows of your data based according to the preferred order in the specified variable.

```{r}
# order the data by ascending length
BirdNest %>% 
  arrange(Length) %>% 
  head()
BirdNest %>% 
  arrange(Length, No.eggs)
```

**Try Out:** Could you output top 10 observations with largest length?

```{r}
BirdNest %>%
  arrange(desc(Length) ) %>% 
  head(n = 10)

```

### Summarise

When you want to create a summary across the data, `summarise()` function in dplyr package can be used. Generally, it often combines with group_by, which creates a summary by subgroups.

```{r}
# to create a summery about average, variance of lengt, and count distinct egg color 
BirdNest %>% 
  summarise(
    mean_length = mean(Length), 
    var_Length = var(Length), 
    n_distict_color = n_distinct(Color)
    ) 

# to create a summery respective to same fields above within each egg color.
BirdNest %>% 
  group_by(Color) %>% 
  summarise(
    mean_length = mean(Length), 
    var_Length = var(Length), 
    n_distict_color = n_distinct(Color)
    ) 

```

### group by

`group_by()` and `summarise()` together can create a split-apply-combine analysis. group_by() splits the data into groups, `summarise()` provides summary function in each group and the summary for each subgroups are combined and returned.

[Note: Adding multiple variables in group_by() will return a summary with grouping by adding order.]{.underline}

```{r}
# create summary about mean and variance of legnth, number of distinct location by color and nesttype

BirdNest %>% 
  group_by(Color,Nesttype) %>% 
  summarise(
    mean_length = mean(Length), 
    num_obs = n(), 
    n_distict_location = n_distinct(Location)
    )
```

**Try Out:**

create a summary about the largest ratio `(ratio = Nestling / Totcare)` by nest type (excluding "`cavity`" category) and present the result in a descending order?

```{r , echo=FALSE}
BirdNest %>% 
  filter(!grepl("cavity",Nesttype)) %>% 
  mutate(ratio = Nestling / Totcare) %>% 
  group_by(Nesttype) %>% 
  summarise(max_ratio = max(ratio, na.rm=T)) %>%
  arrange(desc(max_ratio)) 
```

# Data Visualization

We are going to use the package of ggplot2 based on *The Grammar of Graphics* by [Leland Wilkinson](https://www.cs.uic.edu/~wilkinson/TheGrammarOfGraphics/GOG.html). He is very influential in R programming - he is also the author of the "tidyverse.\
ggplot2 is probably the MOST popular plotting package in R, and heavily influences graphing grammars in other languages, notably python and js.

## Format

1.  start with ggplot function - specify data and "aesthetics"
2.  add "geoms" which are the type of plot (points, lines, bars, etc)
3.  add "scales" which further specify the aesthetics: axes, colors, sizes
